Equalizer 2.0\\Convergence of a Parallel Rendering Framework

### 2017-04-03 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@39 db7c848b-7755-4df2-b763-8796797afa52

%----------------------------------------------------------------------
\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
    untracked views running RTT DeltaGen for virtual car usability studies (c)
    Cave2 running a molecular visualization build based on Omegalib.}
  \label{FIG_teaser}
\end{figure*}

@@

\subsection{New Decomposition Modes}

The initial version of Equalizer implemented sort-first (2D), sort-last (DB),
stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
present new decomposition modes and motivate their use case. \fig{fig:compounds}
provides an overview of the new modes. We describe their configuration in the
\textit{compound} framework introduced in \cite{EMP:09}.

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
  \end{subfigure}\\[\medskipamount]
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source3"
    phase 2 period 3
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}\vspace{24ex}}
    \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source1"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source2"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source3"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
  inputframe { name "frame.source3" }
}
    \end{lstlisting}}
    \caption{\label{fig:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 3 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    pixel [ 1 0 3 1 ]
    outputframe {}
  }
  compound {
    channel "source2"
    pixel [ 2 0 3 1 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 3 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    subpixel [ 1 3 ]
    outputframe {}
  }
  compound {
    channel "source2"
    subpixel [ 1 3 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New Equalizer task decompositions and their compounds for parallel
    rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to Equalizer
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined in \sref{sec:threading}. The output frame
rate of the destination channel is smoothened using a frame rate equalizer
(\sref{sec:framerateEq}).

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks are queued and
processed by all source channels by polling a server-central queue. Prefetching
ensures that the task communication overlaps with rendering. As shown in
\cite{SPEP:16}, these modes provide better performance due to being inherently
load-balanced, as long as there is an insignificant overhead for the render task
setup. This mode is transparent to Equalizer applications.

@@

\subsection{Equalizers}

Equalizer are an addition to compound trees. They modify parameters of their
respective subtree at runtime to optimize one aspect of the decomposition. Due
to their nature, they are transparent to application developers, but might have
application-accessible parameters to tune their behaviour.

@@

We have implemented various tunable parameters allowing application developers
to optimize the load balancing based on the characteristics of their rendering
algorithm:
\begin{compactdesc}
\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
  defining how much of the computed delta from the previous position is
  applied. The equal load distribution within the region of interest assumed by
  the load equalizer is in reality not equal, causing the load balancing to
  overshoot.
\item[Resistance] eliminates small deltas in the load balancing step. This might
  help the application to cache some computations since the frustum does not
  change each frame.
\item[Boundaries] define the modulo factor in pixels onto which a load split may
  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
  position, e.g., screen door transparency, which can be eliminated by aligning
  the boundary to the pixel repetition. Furthermore, some rendering algorithms
  are sensitive to cache alignments, which can again be exploited by chosing the
  corresponding boundary.
\end{compactdesc}

@@

\subsection{Optimizations}

@@

Compression for the compositing step in parallel rendering is critical for
performance. This not only applies to the well-researched network transfer step,
but also for the transfer between GPU and CPU. Equalizer supports a variety of
compression algorithms, from very fast RLE encoding, JPEG compression to YUV
subsampling on the GPU. These algorithms are implemented as runtime-loaded
plugins, allowing easy extension and customization to application-specific
compression. In \cite{MEP:10}, we show this to be a critical step for
interactive performance at scale.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. Equalizer has been used in many virtual reality
installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
at the KAUST visualization laboratory and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current view direction af the
user. Initial experiments show that this is particularly effective for objects
placed within the immersive space, that is, in front of all display segments.

@@

\subsection{Physical and Logical Visualization Setup}

@@

\subsection{Runtime Reconfiguration}\label{sec:reconfig}

@@

\subsection{Automatic Configuration}

@@

\subsection{Qt Windowing}

@@

\subsection{Tide Integration}

@@

\subsection{Sequel}\label{sec:sequel}

@@

\subsection{Architecture}

@@

\subsection{Infiniband RDMA}

@@

\subsection{Reliable Stream Protocol}\label{sec:RSP}

RSP is an implementation of a reliable multicast protocol over unreliable UDP
transport. RSP behaves similarly to TCP, it provides full reliability and
ordering of the data, and slow receivers will eventually throttle the
sender. This behaviour is needed to guarantee delivery of data in all
situations. Pragmatic generic multicast (PGM~\cite{pgm}) provides full ordering,
but slow clients will disconnect from the multicast session instead of
throttling the send rate.

@@

\subsection{Distributed, Versioned Objects}

@@

\subsection{Barriers, Queues and Object Maps}\label{sec:barrier}

>>>>

%----------------------------------------------------------------------
\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
    untracked views running RTT DeltaGen for virtual car usability studies, and (c)
    Cave2 running a molecular visualization build based on Omegalib.}
  \label{FIG_teaser}
\end{figure*}

@@

%------------------------------------------------------------------------------
\subsection{New Decomposition Modes}

The initial version of Equalizer implemented the basic sort-first (2D), sort-last (DB),
stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
present the newly added decomposition modes and motivate their use case. \fig{fig:compounds}
provides an overview of the new modes. We describe how to setup compound
descriptions in the framework in \cite{EMP:09}.

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
  \end{subfigure}\\[\medskipamount]
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source3"
    phase 2 period 3
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}\vspace{24ex}}
    \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source1"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source2"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source3"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
  inputframe { name "frame.source3" }
}
    \end{lstlisting}}
    \caption{\label{fig:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 3 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    pixel [ 1 0 3 1 ]
    outputframe {}
  }
  compound {
    channel "source2"
    pixel [ 2 0 3 1 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 3 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    subpixel [ 1 3 ]
    outputframe {}
  }
  compound {
    channel "source2"
    subpixel [ 1 3 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New Equalizer task decomposition modes and their compound descriptions for parallel rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to Equalizer
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined below in \sref{sec:threading}. The output frame
rate of the destination channel is smoothened using a {\em frame rate equalizer}
(\sref{sec:framerateEq}).

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks are queued and
processed by all source channels by polling a server-central queue. Prefetching
ensures that the task communication overlaps with rendering. As shown in~
\cite{SPEP:16}, these modes provide better performance due to being inherently
load-balanced, as long as there is an insignificant overhead for the render task
setup. This mode is transparent to Equalizer applications.

@@

%------------------------------------------------------------------------------
\subsection{Equalizers}

{\em Equalizers} are an addition to compound trees. They modify parameters of their
respective subtree at runtime to dynamically optimize one aspect of the decomposition. Due
to their nature, they are transparent to application developers, but might have
application-accessible parameters to tune their behavior.
\FIXME{ok, this is a bit tricky to Equalizer as a name for the framework as well as for this, can't we resolve this?}

@@

We have implemented various tunable parameters allowing application developers
to optimize the load balancing based on the characteristics of their rendering
algorithm:

\begin{compactdesc}
\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
  defining how much of the computed delta from the previous position is
  applied. The equal load distribution within the region of interest assumed by
  the load equalizer is in reality not equal, causing the load balancing to
  overshoot.
\item[Resistance] eliminates small deltas in the load balancing step. This might
  help the application to cache some computations since the frustum does not
  change each frame.
\item[Boundaries] define the modulo factor in pixels onto which a load split may
  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
  position, e.g., screen door transparency, which can be eliminated by aligning
  the boundary to the pixel repetition. Furthermore, some rendering algorithms
  are sensitive to cache alignments, which can again be exploited by chosing the
  corresponding boundary.
\end{compactdesc}

@@

%------------------------------------------------------------------------------
\subsection{Optimizations}

@@

Compression for the compositing step in parallel rendering is critical for
performance. This not only applies to the well-researched network transfer step,
but also for the transfer between GPU and CPU. Equalizer supports a variety of
compression algorithms, from very fast RLE encoding, YUV
subsampling on the GPU to JPEG compression. These algorithms are implemented as runtime-loaded
plugins, allowing easy extension and customization to application-specific
compression. In \cite{MEP:10}, we show this to be a critical step for
interactive performance at scale.

@@

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. Equalizer has been used in many virtual reality
installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
at the KAUST visualization laboratory and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations. All features presented were motivated by application use
cases and have been validated with their respective users.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current view direction of the
user. Initial experiments show that this is particularly effective for objects
placed within the immersive space, that is, in front of all display segments.

@@

%------------------------------------------------------------------------------
\subsection{Physical and Logical Visualization Setup}

@@

%------------------------------------------------------------------------------
\subsection{Runtime Reconfiguration}\label{sec:reconfig}

@@

%------------------------------------------------------------------------------
\subsection{Automatic Configuration}

@@

%------------------------------------------------------------------------------
\subsection{Qt Windowing}

@@

%------------------------------------------------------------------------------
\subsection{Tide Integration}

@@

%------------------------------------------------------------------------------
\subsection{Sequel}\label{sec:sequel}

@@

%------------------------------------------------------------------------------
\subsection{Architecture}

@@

%------------------------------------------------------------------------------
\subsection{Infiniband RDMA}

@@

%------------------------------------------------------------------------------
\subsection{Reliable Stream Protocol}\label{sec:RSP}

RSP is an implementation of a reliable multicast protocol over unreliable UDP
transport. RSP behaves similarly to TCP, it provides full reliability and
ordering of the data, and slow receivers will eventually throttle the
sender. This behavior is needed to guarantee delivery of data in all
situations. Pragmatic generic multicast (PGM~\cite{pgm}) provides full ordering,
but slow clients will disconnect from the multicast session instead of
throttling the send rate.

@@

%------------------------------------------------------------------------------
\subsection{Distributed, Versioned Objects}

@@

%------------------------------------------------------------------------------
\subsection{Barriers, Queues and Object Maps}\label{sec:barrier}
